Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-14 paper/sections/proof-reconstruction.tex
[ paper ] citation using natbib.

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by Sultana in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is the proof
assistant, \Agda. We choose \Agda, but a proof assistant with
support of type theory and similar features like \Agda
(see a description of \Agda in section
\ref{sec:implementation}) can be used.
Following the diagram in Fig.
\ref{fig:proof-reconstruction-workflow}, the process can begin with
a problem in \CPL that we encode in \TPTP format to be sent as the
input of the \Metis prover.
The proof-search algorithms of \Metis in step No. 3 intend to find
a solution; if they success, \Metis replies a derivation in \TSTP
format of the proof, else \Metis may no stop and running forever or
replies the problem is counter feasible.

@@

Finally, we use the proof assistant in step No. 7 to type-check the
proof and verify its correctness. Next, we only have one answer:
type-checking succeed (i.e., the proof is valid) or type-checking
failed.
If the type-check fails, one of the three actors in the process are
responsible: \Metis, \Athena, or \Agda. We are responsible for
errors committed in the translation from \TSTP to \Agda; lack of the
coverage of the \Metis' inference rules, or bad assumptions about
them. \improvement{Make this statement stronger and clear.}

@@

\subsection{Well-Structured Recursion}
\label{ssec:well-structured recursion}

In the process to emulate \Metis' inference rules in type theory,
We encountered many recursive algorithms with termination problems,
that is,when an algorithm could run forever or we can not guarantee
that it will stop after some point. Such a problem makes impossible
to translate the algorithm in type theory directly.
For that reason, we show a technique that will help us to define
new algorithms based on the previous ones to solve the problem
mentioned.

A recursive function is a \emph{well-structured recursion}, if the
size of the arguments with which it is defined decrease in
size for each call to the function in the definition~\cite{Bove2005}.
A well-structured function terminates and it can be translated
directly to type theory~\cite{Bove2002}.

Let us consider the following example from a inner function
in the \strip inference.

\begin{equation}
\label{eq:unshunt-definition}
\unshunt(φ) =
\begin{cases}
\unshunt((φ₁ ∧ φ₂) ⇒ φ₃)
  & φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃),\\

\unshunt(φ₁ ⇒ φ₂) ∧ \unshunt(φ₁ ⇒ φ₃)
  & φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃),\\

@@

From the definition above, in the first case of the \unshunt
function, the size of the argument doesn't really decrease.
Actually, it maintains with the same length syntactically
(i.e., (φ₁ ∧ φ₂) ⇒ φ₃ compared with (φ₁ ⇒ (φ₂ ⇒ φ₃))).
Therefore, a termination checking will fail with \unshunt function.

@@

Thus, we have a new function
$\unshunt : \PROP \to \Nat \to \PROP$
defined by structural recursion in its second argument in
Eq.~\ref{eq:unshunt-structured}.

\begin{align}
\label{eq:unshunt-structured}
\begin{split}
\unshunt(φ, \zero) &= φ\\
\unshunt(φ, \suc(n)) &=
\begin{cases}
\unshunt((φ₁ ∧ φ₂) ⇒ φ₃, n)
  & φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃),\\
\unshunt(φ₁ ⇒ φ₂, n)  & φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃).\\
∧\ \unshunt(φ₁ ⇒ φ₃, n)
\end{cases}
\end{split}
\end{align}

To feed the second argument of this new function, we will use
functions. For instance, we opted to use a complexity measure of
the formula that \unshunt function transforms.

@@

Therefore, we define an appropriate complexity measure
$\unshunt_{cm}: \PROP → \NN$
for the second argument of the $\unshunt$ function in
Eq.~\ref{eq:unshunt-rank-definition}.

@@

Then, a curated version of the \unshunt functions is presented in Eq.
\ref{eq:unshunt-fixed}.

\begin{equation}
\label{eq:unshunt-fixed}
\unshunt(\varphi) := \unshunt(\varphi, \unshunt_{cm}(\varphi)).
\end{equation}

@@

We reformulate \mysplit function to match with the
new definition after apply the process described above in
subsection~\ref{ssec:well-structured recursion}, since
in the recursion calls of \mysplit in Eq.~\ref{eq:split-definition},
the negations, $(¬ φ_{1,2})$ raise a terminating problem.

>>>>

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by Sultana et al in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is the proof
assistant, \Agda. We choose \Agda, but a proof assistant with
support of type theory and similar features like \Agda
(see a description of \Agda in section
\ref{sec:implementation}) can be used.
Following the diagram in Fig.
\ref{fig:proof-reconstruction-workflow}, the process can begin with
a problem in \CPL that we encode in \TPTP format to be sent as the
input of the \Metis prover.
The algorithms of \Metis in step~3 intend to find
a solution; if they success, \Metis replies a derivation in \TSTP
format of the proof, else the problem is counter feasible.

@@

Finally, we use the proof assistant in step 7 to type-check the
proof and verify its correctness. Next, we only have one answer:
type-checking succeed (i.e., the proof is valid) or type-checking
failed.
If the type-check fails, one of the three actors in the process are
responsible: \Metis, \Athena, or \Agda. We are responsible for
errors committed in the translation from \TSTP to \Agda; lack of the
coverage of the \Metis' inference rules, or bad assumptions about
them. \improvement{Make this statement stronger and clear.}

@@

\subsection{Small Structured Recursion}
\label{ssec:small-structured-recursion}

In the process to emulate \Metis' inference rules in type theory,
we encountered many recursive algorithms with termination problems,
that is,when an algorithm could run forever or we can not guarantee
that it will stop after some point. Such a problem makes impossible
to translate the algorithm in type theory directly.
For that reason, we show a technique that will help us to define
new algorithms based on the previous ones to solve the problem
mentioned.

A recursive function is a \emph{small-structured-recursion}, if the
size of the arguments with which it is defined decrease in
size for each call to the function in the definition~\cite{Bove2005}.
A well-structured function terminates and it can be translated
directly to type theory~\cite{Bove2002}.

Let us consider the following example from an inner function
in the \strip inference.

\begin{equation}
\label{eq:unshunt-definition}
\unshunt(φ)_{0} =
\begin{cases}
\unshunt_{0}((φ₁ ∧ φ₂) ⇒ φ₃)
  & φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃),\\

\unshunt_{0}(φ₁ ⇒ φ₂) ∧ \unshunt_{0}(φ₁ ⇒ φ₃)
  & φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃),\\

@@

From the definition above, in the first case of the \unshunt
function, the size of the argument in the recursion,(φ₁ ∧ φ₂) ⇒ φ₃,
is not structured smaller than the decrease, (φ₁ ⇒ (φ₂ ⇒ φ₃)).
Therefore, a termination checking will fail with the \unshunt
function.

@@

Thus, we have a new function
$\unshunt_{1} : \PROP \to \Nat \to \PROP$
defined by structural recursion in its second argument in
Eq.~\ref{eq:unshunt-structured}.

\begin{align}
\label{eq:unshunt-structured}
\begin{split}
\unshunt_{1}(φ, \zero) &= φ\\
\unshunt_{1}(φ, \suc(n)) &=
\begin{cases}
\unshunt_{1}((φ₁ ∧ φ₂) ⇒ φ₃, n)
  & φ ≡ φ₁ ⇒ (φ₂ ⇒ φ₃),\\
\unshunt_{1}(φ₁ ⇒ φ₂, n)  & φ ≡ φ₁ ⇒ (φ₂ ∧ φ₃).\\
∧\ \unshunt_{1}(φ₁ ⇒ φ₃, n)
\end{cases}
\end{split}
\end{align}

To feed the second argument of this new function, we will use
functions. For instance, we opted to use a complexity measure of
the formula that $\unshunt_{1}$ function transforms.

@@

Therefore, we define an appropriate complexity measure
$\unshunt_{cm}: \PROP → \NN$
for the second argument of the $\unshunt_{1}$ function in
Eq.~\ref{eq:unshunt-rank-definition}.

@@

Then, a curated version of the $\unshunt_{0}$ function is presented in Eq.
\ref{eq:unshunt-fixed}.

\begin{equation}
\label{eq:unshunt-fixed}
\unshunt(\varphi) := \unshunt_{1}(\varphi, \unshunt_{cm}(\varphi)).
\end{equation}
\change{changes the name of the functions to avoid different
functions with the same name.}

@@

We reformulate \mysplit function to match with the
new definition after apply the process described above in
subsection~\ref{ssec:small-structured-recursion}, since
in the recursion calls of \mysplit in Eq.~\ref{eq:split-definition},
the negations, $(¬ φ_{1,2})$ raise a terminating problem.
